---
title: "<img class='image1' src='images/sgplot-logo.svg' alt=''/>Creating accessible charts in R using sgplot"
author: "Alice Byers"
institute: "Data Innovation Team, Scottish Government"
date: "19 June 2023"
format:
  revealjs: 
    embed-resources: true
    slide-number: true
    logo: images/SG-logo.jpg
    controls: true
    theme: [default, styles.scss]
execute:
  echo: TRUE
  eval: TRUE
lang: en
---

## Aims

* Explain why analysts should use sgplot

*	Demonstrate how to use sgplot

*	Offer some general guidance on creating good charts for Scottish Government analysis


## Why use sgplot


## Installation


## Example without sgplot (default ggplot2 style)

```{r}
#| code-line-numbers: "1-4|6-22"
 
# Load packages
library(ggplot2)
library(dplyr)
library(gapminder)

# Create line chart
gapminder |>
  filter(country %in% c("United Kingdom", "China")) |>
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 82),
                     breaks = seq(0, 80, 20),
                     expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
  labs(
    x = "Year",
    y = NULL,
    title = "Living Longer",
    subtitle = "Life Expectancy in the United Kingdom and China 1952-2007",
    caption = "Source: Gapminder",
    colour = NULL
  )
```

## use_sgplot()

```{r}
#| code-line-numbers: "1|3"
 
library(sgplot)

use_sgplot()
```


## Package website

<https://datasciencescotland.github.io/sgplot>{preview-link="true"}

* Cookbook

* Function reference

* Further guidance on accessibility, colours and saving charts

* Guidance on how to ask for help or contribute


## Future developments

* Highcharter

* Anything that would make your life easier!

  * If your team has its own branding style and/or colour palettes, these can be added to sgplot (as long as they're accessible!)
  
  
## Links

* [sgplot website](https://datasciencescotland.github.io/sgplot)

* [GitHub repository](https://github.com/DataScienceScotland/sgplot/)

* [Contributing guidance](https://datasciencescotland.github.io/sgplot/CONTRIBUTING.html)

* Contact me at <alice.byers@gov.scot>