---
title: "Accessible data visualisations"
subtitle: "Scottish Government Accessibility Champions Network"
author: "Alice Byers"
institute: "Data Innovation Team, Data Division"
date: "19 October 2023"
format:
  revealjs: 
    embed-resources: true
    slide-number: true
    logo: images/SG-logo.jpg
    controls: true
    link-external-newwindow: true
    theme: [default, styles.scss]
execute:
  echo: false
  fig-format: "svg"
lang: en
editor_options: 
  chunk_output_type: console
---

## About me

* [Reproducible Analytical Pipeline (RAP)](https://analysisfunction.civilservice.gov.uk/support/reproducible-analytical-pipelines/) developer

* Background in statistics and data analysis

* New to the Accessibility Champions Network

::: {.notes}
* Work with statisticians to improve their analytical processes and outputs
* Use R programming language to write code for data cleaning, analysis, visualisation and reporting.
* Worked on a project last year to support a team to develop a dashboard where I learned a lot about accessibility and how to make data visualisations accessible.
* Not an expert!
:::


## Agenda

* Examples of charts and common issues

* An introduction to the [sgplot R package](https://datasciencescotland.github.io/sgplot/)

* An overview of available guidance for making charts accessible


# Example {.h1-blue}

```{r run-examples}
if(!all(c(paste0("ex", 1:3, ".svg"), "table.html") %in%
        list.files("images/"))) {
  source("examples.R")
}
```

```{r alt-text}

ex1_alt <- paste(
  "A multiple line chart with year on x axis, life expectancy on y axis and",
  "separate countries with colour. The chart background is grey with white",
  "grid lines, and the data lines are turquoise and orange."
)

ex2_alt <- paste(
  "A multiple line chart with year on x axis, life expectancy on y axis and",
  "separate countries with colour. The chart background is grey with white",
  "grid lines, and the data lines are turquoise and orange."
)

ex3_alt <- paste(
  "The life expectancy for both the United Kingdom and China have increased",
  "between 1952 and 2007. The United Kingdom has consistently had a higher",
  "life expectancy than China throughout this period."
)
```

## Example 1 {.smaller auto-animate=true}

![](images/ex1.svg){fig-alt="`r ex1_alt`"}

[Slido poll](https://app.sli.do/event/pZMm3JwX6jxNQgE8Mn7FAu/live/polls){preview-link="true"}: What could be improved?


## Example 1 {auto-animate=true}

![](images/ex1.svg){fig-alt="`r ex1_alt`" .absolute top=25 right=0 width=500 height=250}

<br>
<br>
<br>

::: {.incremental}

* Cluttered; lots of grid lines, lots of colours, lots of data

* Poor colour contrast with background and with other colours

* Angled text not easy to read

:::

::: {.notes}
* Clutter
    * Difficult to pick out important features
    * Overwhelming

* Text
    * Angled text means you have to tilt your head to read
    * Labels don't make sense
:::


# Colour {.h1-blue}


## Things to think about

<br>

:::: {.columns}

::: {.column width="60%"}
::: {.r-stack}
![](images/ex1.svg){fig-alt="`r ex1_alt`"}

![](images/ex1_bw.svg){fig-alt="`r ex1_alt`" .fragment fragment-index=4}
:::
:::

::: {.column width="40%"}

::: {.fragment fragment-index=1}
* Contrast with background
:::

::: {.fragment fragment-index=2}
* Contrast with other elements
:::

::: {.fragment fragment-index=3}
* Reliance on colour
:::

:::

::::


## Colour palettes {.smaller}

* Scottish Government Design System [colour palettes for data visualisations](https://designsystem.gov.scot/guidance/charts/data-visualisation-colour-palettes)

* Example of palette for categorical data:

```{r palettes}
sgplot:::colour_table(sgplot::sg_colour_palettes$main)
```

<br style="line-height: 5px"/>

* Other palettes available for sequential and focus charts (with caveats)

::: {.notes}
* All 4 colours have sufficient contrast (3:1) with a white background and adjacent colours in the palette.
:::


# Example {.h1-blue}

## Example 2 {.smaller auto-animate=true}

![](images/ex2.svg){fig-alt="`r ex2_alt`"}

. . .

[Slido poll](https://app.sli.do/event/pZMm3JwX6jxNQgE8Mn7FAu/live/polls){preview-link="true"}: What could be improved?

::: {.notes}
* Less data
* Fewer grid lines
* All text horizontal
* Accessible colour palette
* However, there are still issues
:::


## Example 2 {auto-animate=true}

![](images/ex2.svg){fig-alt="`r ex2_alt`" .absolute top=25 right=0 width=500 height=250}

<br>
<br>
<br>

::: {.incremental}

* No way to link data lines to legend without colour

* Meaningless axis label

* No title, subtitle or source

:::

::: {.notes}

:::


## Example 3

### Living Longer
#### Life Expectancy in the United Kingdom and China 1952-2007

![](images/ex3.svg){fig-alt="`r ex3_alt`"}
<div class="source">Source: Gapminder</div>

::: {.notes}
* Much improved compared to original chart
* Almost definitely still improvements that could be made
:::


# Alternative Text {.h1-blue}


## Forms of alt text

* Commonly sits in the code behind the chart to be read by screen reader software

::: {.incremental}

* However, it can also be:

    * a text description in the main body of the page
    
    * a table of the data presented in the chart

:::


## Text description

<br>

`r ex3_alt`


## Data table

```{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string('images/table.html')
```
```

# sgplot {.h1-blue}


## sgplot

![](images/sgplot-logo.svg){fig-alt="" .absolute top=25 right=0}

* An [R](https://www.r-project.org/) package to apply a style and colour palettes to<br>[`ggplot2`](https://ggplot2.tidyverse.org/) charts

* Applies [Scottish Government Design System colour palettes](https://designsystem.gov.scot/guidance/charts/data-visualisation-colour-palettes) and accessibility best practice

* [Package website](https://datasciencescotland.github.io/sgplot/) with documentation, guidance and examples

* Open project with [code publicly available on GitHub](https://github.com/DataScienceScotland/sgplot/)


## Example - `ggplot2` defaults {.smaller}

```{r ggplot2-example, echo = TRUE}
#| code-fold: true

library(ggplot2)

dat <- data.frame(
  shop = rep(c("Shop 1", "Shop 2"), each = 2),
  fruit = rep(c("Apples", "Bananas"), times = 2),
  number = c(5, 4, 6, 2)
)

ggplot(dat) +
  geom_col(aes(x = shop, y = number, fill = fruit),
           position = "dodge")
```


## Example - `sgplot` defaults {.smaller}

```{r sgplot-example, echo = TRUE}
#| code-fold: true

library(sgplot)

use_sgplot()

ggplot(dat) +
  geom_col(aes(x = shop, y = number, fill = fruit),
           position = "dodge")
```

::: {.notes}
* There are obviously still things to improve about this chart; e.g. labels, linking of colour from chart to legend
* However, it's a big improvement on the original chart with very little effort
:::


## Future Developments

* Support for interactive chart packages

* Use Roboto Google font as per Scottish Government design system

* Anything that would make your life easier!

    * If your team has its own branding style and/or colour palettes, these can be added to sgplot (as long as they’re accessible!)


## Get Involved

* [Open to feedback](https://github.com/DataScienceScotland/sgplot/issues)

* [Open for contributions](https://datasciencescotland.github.io/sgplot/CONTRIBUTING.html) (however big or small!)

* Become a package maintainer


# Guidance & Resources {.h1-blue}

## Scottish Government

* [Data visualisation accessibility guidance](https://designsystem.gov.scot/guidance/charts)

* [Data visualisation colour palettes](https://designsystem.gov.scot/guidance/charts/data-visualisation-colour-palettes)

* [sgplot guidance](https://datasciencescotland.github.io/sgplot/articles/)

* Viva Engage (Yammer) communities

    * [Digital Accessibility](https://web.yammer.com/main/groups/eyJfdHlwZSI6Ikdyb3VwIiwiaWQiOiI1Mjc2NzA0NzY5In0/all)

    * [Communicating Analysis with Impact](https://web.yammer.com/main/groups/eyJfdHlwZSI6Ikdyb3VwIiwiaWQiOiI1NjA2MTA1MDg5In0/all)


## Analysis Function

* [Charts](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-charts/)

* [Colours](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/)

* [Making analytical publications accessible](https://analysisfunction.civilservice.gov.uk/policy-store/making-analytical-publications-accessible/)

* [Dashboards](https://analysisfunction.civilservice.gov.uk/policy-store/top-tips-for-designing-dashboards/)

* [Data visualisation e-learning](https://analysisfunction.civilservice.gov.uk/support/communicating-analysis/introduction-to-data-visualisation-e-learning/)


## Contact

<br>
<alice.byers@gov.scot>

