---
title: "Accessible data visualisations"
subtitle: "Science and Engineering in Government Conference"
author: "Alice Hannah"
institute: "Data Innovation Team, Scottish Government"
date: "12 November 2024"
format:
  revealjs: 
    embed-resources: true
    slide-number: true
    logo: images/SG-logo.jpg
    controls: true
    link-external-newwindow: true
    preview-links: true
    theme: [default, styles.scss]
execute:
  echo: false
  fig-format: "svg"
lang: en
editor_options: 
  chunk_output_type: console
---

## Agenda

* Examples of charts and common accessibility issues

* How to use colour

* How to provide alternative text

* An overview of available guidance


# Example {.h1-blue}

```{r run-examples}
if(!all(c(paste0("ex", 1:3, ".svg"), "table.html") %in%
        list.files("images/"))) {
  source("examples.R")
}
```

```{r alt-text}

ex1_alt <- paste(
  "A line chart showing a time trend of life expectancy for different",
  "countries. The background is light grey with lots of white grid lines.",
  "There are six trend lines differentiated by colour with a legend to the",
  "right hand side of the chart. The y axis title is at a 90 degree angle and",
  "the x axis labels are at a 45 degree angle."
)

ex1_bw_alt <- paste(
  "A line chart showing a time trend of life expectancy for different",
  "countries. The background is light grey with lots of white grid lines.",
  "There are six trend lines with a legend to the right hand side of the",
  "chart. All lines are black. The y axis title is at a 90 degree angle and",
  "the x axis labels are at a 45 degree angle."
)

ex2_alt <- paste(
  "A line chart showing a time trend of life expectancy for different",
  "countries. The background is white with light grey, horizontal grid lines.",
  "There are two trend lines, one dark blue and one turquoise, with a legend",
  "to the right hand side of the chart. All text is horizontal."
)

ex3_alt <- paste(
  "A line chart showing a time trend of life expectancy for different",
  "countries. The background is white with light grey, horizontal grid lines.",
  "There are two trend lines, one dark blue and one turquoise. The lines are",
  "directly labelled with the country they represent. All text is horizontal."
)

ex3_alt_story <- paste(
  "The life expectancy for both the United Kingdom and China have increased",
  "between 1952 and 2007. The United Kingdom has consistently had a higher",
  "life expectancy than China throughout this period."
)
```

## Example 1 {.smaller auto-animate=true}

![](images/ex1.svg){fig-alt="`r ex1_alt`"}

In the Teams chat: [What could be improved?]{style="color:#0065bd;font-weight: bold;"}

::: {.notes}
* No context given
:::


## Example 1 {auto-animate=true}

![](images/ex1.svg){fig-alt="`r ex1_alt`" .absolute top=25 right=0 width=500 height=250}

<br>
<br>
<br>

::: {.incremental}

* Cluttered; lots of grid lines, lots of colours, lots of data

* Poor colour contrast with background and with other colours

* Text is quite small and difficult to read at an angle

:::

::: {.notes}
* Clutter
    * What is the chart trying to tell us?
    * Overwhelming

* Colour
    * Some colours are difficult to tell apart
    * Could cause problems for people with low vision or cognitive differences

* Text
    * Angled text means you have to tilt your head to read
    * Labels don't make sense
:::


# Colour {.h1-blue}

::: {.notes}
* Colour can be a very powerful tool to highlight elements of a chart and make it looks visually appealing.

* However, it is also very easy to use it badly which can confuse and exclude some users.
:::


## Things to think about

<br>

:::: {.columns}

::: {.column width="65%"}
::: {.r-stack}
![](images/ex1.svg){fig-alt="`r ex1_alt`"}

![](images/ex1_bw.svg){fig-alt="`r ex1_bw_alt`" .fragment fragment-index=4}
:::
:::

::: {.column width="35%"}

::: {.fragment fragment-index=1}
* Contrast with background
:::

::: {.fragment fragment-index=2}
* Contrast with other elements
:::

::: {.fragment fragment-index=3}
* Reliance on colour
:::

:::

::::


::: {.notes}
* Colour contrast is a measure of how different two colours are from each other.

* Graphical elements must have a 3:1 colour contrast ratio with the background and with adjacent colours.

* A good way to check whether you're relying on colour is to remove it. Now colour has been removed, we can probably say that for these countries, there's been a general increase over the time period. But there's one country that has dropped off, and without colour, we have no way of knowing which one this is.
    * We'll come back to this
:::


## Colour palettes {.smaller}

* Scottish Government Design System [colour palettes for data visualisations](https://designsystem.gov.scot/guidance/charts/data-visualisation-colour-palettes){preview-link="false"}

* Example of palette for categorical data:

```{r palettes}
sgplot:::colour_table(sgplot::sg_colour_palettes$main) |>
  gt::tab_options(table.font.size = gt::pct(80),
                  table.width = gt::pct(75))
```

* Palettes for sequential and focus charts are also available

::: {.notes}
* All colours have sufficient contrast (3:1) with a white background and adjacent colours in the palette. Therefore, it's important to use them in this order.

* Recommend using a maximum of 4 colours at a time to avoid clutter. However, there is also an extended palette available if absolutely necessary.
:::


# Example {.h1-blue}

## Example 2 {.smaller auto-animate=true}

![](images/ex2.svg){fig-alt="`r ex2_alt`"}

. . .

In the Teams chat: [What could be improved?]{style="color:#0065bd;font-weight: bold;"}

::: {.notes}
* Some improvements
    * Less data
    * Fewer grid lines
    * All text horizontal
    * Accessible colour palette

* However, there are still issues
:::


## Example 2 {auto-animate=true}

![](images/ex2.svg){fig-alt="`r ex2_alt`" .absolute top=25 right=0 width=500 height=250}

<br>
<br>
<br>

::: {.incremental}

* No way to link data lines to legend without colour

* Meaningless axis label

* No title, subtitle or source

:::

::: {.notes}
* Fewer colours used and contrast requirements are met
    * However, you still can't link the data to the legend without use of colour
    
* The axis labels are now horizontal, but they still don't make sense.

* We still don't have the context for this chart. What is the data, why are we seeing this chart? What is the story?
:::


## Example 3

### Living Longer
#### Life Expectancy in the United Kingdom and China 1952-2007

![](images/ex3.svg){fig-alt="`r ex3_alt`"}
<div class="source">Source: Gapminder</div>

::: {.notes}
* Much improved compared to original chart
    * Directly labelled lines to avoid reliance on colour
    * Given the y axis a proper title
    * Added some context; headline title, subtitle and source
    
* Almost definitely still improvements that could be made
:::


# Alternative Text {.h1-blue}

::: {.notes}
* We've focussed so far on visual elements of a chart, but what if you're not able to access the visual elements?
:::


## Forms of alt text

* Commonly sits in the source code behind the chart to be read by screen reader software

::: {.incremental}

* However, it can also be:

    * a text description in the main body of the page
    
    * a table of the data presented in the chart

:::

::: {.notes}
* Which form you use should depend on what you expect people to get from the chart.
    * A text description is better for conveying an overall 'story' of the chart
    * A table is better if it's important for users to read the individual data points
    * You can also use both!
    
* The best text description of a chart may also depend on the context. For example, the alt text for the example charts in these slides so far talk about the visual elements of the chart; e.g. what colours are used, how many lines there are, where the legend is positioned, etc. 

* However if these charts were in a statistical publication about life expectancy, your alt text would probably be something more like...
:::


## Example - text description

<br>

`r ex3_alt_story`


## Example - data table

```{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string('images/table.html')
```
```

::: {.notes}
* If using a table, it should be produced in html, rather than an image of a table. As this will not be readable by a screen reader.
:::


# Guidance & Resources {.h1-blue}

## Scottish Government

* [Data visualisation accessibility guidance](https://designsystem.gov.scot/guidance/charts){preview-link="false"}

* [Data visualisation colour palettes](https://designsystem.gov.scot/guidance/charts/data-visualisation-colour-palettes){preview-link="false"}

* [sgplot R package](https://scotgovanalysis.github.io/sgplot/)

* Viva Engage communities (internal only)

    * [Digital Accessibility](https://web.yammer.com/main/groups/eyJfdHlwZSI6Ikdyb3VwIiwiaWQiOiI1Mjc2NzA0NzY5In0/all)

    * [Communicating Analysis with Impact](https://web.yammer.com/main/groups/eyJfdHlwZSI6Ikdyb3VwIiwiaWQiOiI1NjA2MTA1MDg5In0/all)


## Government Analysis Function

* [Charts](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-charts/)

* [Colours](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/)

* [Making analytical publications accessible](https://analysisfunction.civilservice.gov.uk/policy-store/making-analytical-publications-accessible/)

* [Dashboards](https://analysisfunction.civilservice.gov.uk/policy-store/top-tips-for-designing-dashboards/)

* [Data visualisation e-learning](https://analysisfunction.civilservice.gov.uk/support/communicating-analysis/introduction-to-data-visualisation-e-learning/)

* [Basecamp communities](https://analysisfunction.civilservice.gov.uk/support/communicating-analysis/#basecamp)


## Contact

<br>

<alice.hannah@gov.scot>

