---
title: "sgplot"
subtitle: "An R package for accessible data visualisations in the Scottish Government"
author: "Alice Hannah"
email: "alice.hannah@gov.scot"
lang: en
format:
  revealjs: 
    embed-resources: true
    logo: images/qr.svg
    theme: [default, styles.scss]
    template-partials:
      - title-slide.html
    width: 1280
    height: 720
execute:
  echo: false
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(gapminder)
library(dplyr)
library(ggplot2)
library(sgplot)

life_exp <- 
  gapminder %>%
  select(country, year, lifeExp) %>%
  filter(country %in% c("United Kingdom", "China"))

create_chart <- function() {
  ggplot(life_exp) +
  geom_line(aes(x = year, y = lifeExp, colour = country), linewidth = 1) +
  labs(
    x = "Year",
    y = "Life Expectancy",
    colour = "Country"
  )
}
```

## {#default-chart data-menu-title="Default ggplot2 chart" .center}

```{r chart-default}
#| fig.alt = "A multiple line chart showing life expectancy over time in China and the United Kingdom. The background is grey, with many white grid lines. Labels have a small font size and the y-axis label is angled. The two data lines are coloured in light blue and pink-orange."

create_chart()
```


## {#use-sgplot data-menu-title="use_sgplot()" .center}

::: {.center-h}
`sgplot::use_sgplot()`
:::


## {#sgplot-chart data-menu-title="sgplot chart" .center}

```{r chart-sgplot}
#| fig.alt = "A multiple line chart showing life expectancy over time in China and the United Kingdom. The background is white, with minimal horizontal grid lines in a light grey colour. All text is horizontal and has a large font size. The two data lines are coloured in turquoise and navy blue."

sgplot::use_sgplot(base_size = 16)

create_chart() +
  labs(y = "Life\nExpectancy")
```


## {#website data-menu-title="Website" .center}

::: {.center-h}
![](images/home.svg){fig-alt="A screenshot of the sgplot website homepage: https://scotgovanalysis.github.io/sqplot. Content includes instructions to install the sgplot package, and links to a function reference page, and articles."}
:::


## {#cookbook data-menu-title="Cookbook" .center}

::: {.center-h}
![](images/cookbook.svg){fig-alt="A screenshot of the sgplot website cookbook: https://scotgovanalysis.github.io/sqplot/articles/cookbook. The table of contents lists the chart types included in the cookbook: line charts, bar charts, histograms, scatterplots, and more."}
:::
